Imputation of Incomplete Multilevel Data with \pkg{mice}

### 2022-04-01 Manuscript/Imputation_of_Incomplete_Multilevel_Data.Rmd
Estimate multilevel model after imputation

@@

~~In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.~~ Contemporary decision making is driven increasingly often and sometimes even entirely dictated by data. Although data relevant for decision makers were initially collected using small and well-designed studies, there has been a growing need to address more complex questions at a wider scale. This not only requires to collect larger amounts of data, but also from more diverse settings and populations. 
For example, [TD: introduce IPDMA here; eg. HIV studies Johanna is referring to]. Another example is the use of large databases with information from thousands or even millions of individuals from multiple cities, regions or even countries. Finally, a third situation arises when data are collected from multicenter studies [TD: maybe refer to the schools/classes example here]. For example, students may be clustered in classes in psychometrics research, or patients may be clustered in studies in individual patient data meta-analyses in medical research. A common characteristic in aforementioned examples is the presence of clustering. [TD: briefly explain what is clustering?]

@@

For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow. With the `popmis` data, we show how (and how not) to develop an imputation model. With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods. With the `impact` data we provide an example of multivariate missingness in real-world data. Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.^[TODO: Add notation paragraph or 'translation table' linking multilevel equations to \pkg{lme4} formulas. Use betas instead of gamma's and mu's. Add interpretation of values in predictormatrix (-2 for the cluster variable, 2 for random effects).]

@@

```{r}
set.seed(2022)
library(mice)
library(ggmice)
library(ggplot2)
library(dplyr)
library(lme4)
```

@@

@@

This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` ($t = $ `r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")`).

@@

@@

Create a predictor matrix and remove `school` as predictor for `popular`:

```{r pop_ignored, echo=TRUE, message=FALSE, warning=FALSE}
# predictor matrix
pred <- quickpred(popmis)
plot_pred(pred)

# impute the data, ignoring the cluster structure
imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
```

>>>>

<!-- Comments Thomas: -->
<!-- .	We need to remind the reader about general knowledge about missing data: it usually leads to loss of power and bias; we distinguish between MCAR, MAR and MNAR in theory but in practice this distinction is less clear; multiple imputation is generally recommended to avoid bias (if data are MCAR or MAR, but also works reasonable in some situations where data are partially MNAR) -->
<!-- .	We need to clarify what is clustered data, and provide common types (IPDMA, registries with data from multiple hospitals/regions, and multicenter studies). You can directly refer to the examples you will use in the paper here, and build on the text you already had. -->
<!-- .	We need to clarify why clustering is relevant during imputation, and why this exposes the need for specialized imputation methods and more attention during their implemention ("though shall not simply run mice() on any incomplete dataset") -->
<!-- .	Maybe we should also emphasise that in more modern types of big data sources (E.g. EHR data ), measurement of data is often selective and MAR assumptions are less tenable.  -->

@@

~~In many contemporary data analysis efforts, some form of hierarchical or clustered structure is recorded.~~ Contemporary decision-making is  increasingly often data-driven and sometimes even entirely dictated by data. Although data relevant for decision makers were initially collected using small and well-designed studies, there has been a growing need to address more complex questions at a wider scale. This not only requires to collect larger amounts of data, but also from more diverse settings and populations. 
For example, [TD: introduce IPDMA here; eg. HIV studies Johanna is referring to]. Another example is the use of large databases with information from thousands or even millions of individuals from multiple cities, regions or even countries. Finally, a third situation arises when data are collected from multicenter studies [TD: maybe refer to the schools/classes example here]. For example, students may be clustered in classes in psychometrics research, or patients may be clustered in studies in individual patient data meta-analyses in medical research. A common characteristic in aforementioned examples is the presence of clustering. [TD: briefly explain what is clustering?]

@@

For each of these datasets, we will discuss the nature of the missingness, choose one or more imputation models and evaluate the imputed data, but we will also highlight one specific aspect of the imputation workflow. With the `popmis` data, we show how (and how not) to develop an imputation model. With the `hiv` data we focus on extending the imputation model to include Heckman-type selection-inclusion methods. With the `impact` data we provide an example of multivariate missingness in real-world data. Together, this should give enough scaffolding for applied researchers who are faced with incomplete multilevel data.^[TODO: Add notation paragraph or 'translation table' linking multilevel equations to \pkg{lme4} formulas. Use betas instead of gamma's and mu's. Add interpretation of values in predictormatrix (-2 for the cluster variable, 2 for random effects). Add ICC and congeneality here as well. And make missingness mechanism table as well.]

@@

```{r}
set.seed(2022)
library(mice)
library(ggmice)
library(ggplot2)
library(dplyr)
library(lme4)
library(mitml)
```

@@

TODO: think about what is a meaningful rule of thumb to signal that the user should be worried?

@@

This shows us that there are no apparent differences in the distribution of `texp` depending on the missingness indicator of `popular` (t = `r t.test(popmis$texp ~ is.na(popmis$popular)) %>% broom::tidy(.) %>% .[, c("statistic", "p.value")] %>% round(., 3) %>% unlist() %>% paste(collapse = ",  p = ")`).

@@

```{r pop-cca, echo=FALSE}
est_cca <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popmis)

@@

Create a methods vector and predictor matrix for `popular`, and make sure `school` is not included as predictor:

```{r pop-ignored-pred, echo=TRUE, message=FALSE, warning=FALSE}
meth <- make.method(popmis) # methods vector
pred <- quickpred(popmis)   # predictor matrix
plot_pred(pred)
```

Impute the data, ignoring the cluster structure:
```{r pop-ignored-imp, echo=TRUE, message=FALSE, warning=FALSE}
imp_ignored <- mice(popmis, maxit = 1, pred = pred, print = FALSE)
```

Analyze the imputations:
```{r pop-ignored-fit, echo=TRUE, message=FALSE, warning=FALSE}
fit_ignored <- imp_ignored %>% 
  with(lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school))) 
fit_ignored %>% 
  pool() %>% 
  broom.mixed::tidy()
testEstimates(as.mitml.result(fit_ignored), var.comp = TRUE)
```

### 2022-04-01 Manuscript/Imputation_of_Incomplete_Multilevel_Data.tex
Estimate multilevel model after imputation

\sout{In many contemporary data analysis efforts, some form of
hierarchical or clustered structure is recorded.} Contemporary decision
making is driven increasingly often and sometimes even entirely dictated
by data. Although data relevant for decision makers were initially
collected using small and well-designed studies, there has been a
growing need to address more complex questions at a wider scale. This
not only requires to collect larger amounts of data, but also from more
diverse settings and populations. For example, {[}TD: introduce IPDMA
here; eg. HIV studies Johanna is referring to{]}. Another example is the
use of large databases with information from thousands or even millions
of individuals from multiple cities, regions or even countries. Finally,
a third situation arises when data are collected from multicenter
studies {[}TD: maybe refer to the schools/classes example here{]}. For
example, students may be clustered in classes in psychometrics research,
or patients may be clustered in studies in individual patient data
meta-analyses in medical research. A common characteristic in
aforementioned examples is the presence of clustering. {[}TD: briefly
explain what is clustering?{]}

@@

For each of these datasets, we will discuss the nature of the
missingness, choose one or more imputation models and evaluate the
imputed data, but we will also highlight one specific aspect of the
imputation workflow. With the \texttt{popmis} data, we show how (and how
not) to develop an imputation model. With the \texttt{hiv} data we focus
on extending the imputation model to include Heckman-type
selection-inclusion methods. With the \texttt{impact} data we provide an
example of multivariate missingness in real-world data. Together, this
should give enough scaffolding for applied researchers who are faced
with incomplete multilevel data.\footnote{TODO: Add notation paragraph
  or `translation table' linking multilevel equations to \pkg{lme4}
  formulas. Use betas instead of gamma's and mu's. Add interpretation of
  values in predictormatrix (-2 for the cluster variable, 2 for random
  effects).}

@@

This shows us that there are no apparent differences in the distribution
of \texttt{texp} depending on the missingness indicator of
\texttt{popular} (\$t = \$ -0.873, p = 0.383).

@@

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop_ignored-1} \end{center}

>>>>

\sout{In many contemporary data analysis efforts, some form of
hierarchical or clustered structure is recorded.} Contemporary
decision-making is increasingly often data-driven and sometimes even
entirely dictated by data. Although data relevant for decision makers
were initially collected using small and well-designed studies, there
has been a growing need to address more complex questions at a wider
scale. This not only requires to collect larger amounts of data, but
also from more diverse settings and populations. For example, {[}TD:
introduce IPDMA here; eg. HIV studies Johanna is referring to{]}.
Another example is the use of large databases with information from
thousands or even millions of individuals from multiple cities, regions
or even countries. Finally, a third situation arises when data are
collected from multicenter studies {[}TD: maybe refer to the
schools/classes example here{]}. For example, students may be clustered
in classes in psychometrics research, or patients may be clustered in
studies in individual patient data meta-analyses in medical research. A
common characteristic in aforementioned examples is the presence of
clustering. {[}TD: briefly explain what is clustering?{]}

@@

For each of these datasets, we will discuss the nature of the
missingness, choose one or more imputation models and evaluate the
imputed data, but we will also highlight one specific aspect of the
imputation workflow. With the \texttt{popmis} data, we show how (and how
not) to develop an imputation model. With the \texttt{hiv} data we focus
on extending the imputation model to include Heckman-type
selection-inclusion methods. With the \texttt{impact} data we provide an
example of multivariate missingness in real-world data. Together, this
should give enough scaffolding for applied researchers who are faced
with incomplete multilevel data.\footnote{TODO: Add notation paragraph
  or `translation table' linking multilevel equations to \pkg{lme4}
  formulas. Use betas instead of gamma's and mu's. Add interpretation of
  values in predictormatrix (-2 for the cluster variable, 2 for random
  effects). Add ICC and congeneality here as well. And make missingness
  mechanism table as well.}

@@

TODO: think about what is a meaningful rule of thumb to signal that the
user should be worried?

This shows us that there are no apparent differences in the distribution
of \texttt{texp} depending on the missingness indicator of
\texttt{popular} (t = -0.873, p = 0.383).

@@

\begin{CodeChunk}
\begin{CodeInput}
R> est_cca <- lme4::lmer(popular ~ 1 + sex + texp + sex:texp + (1 + sex | school), popmis) 
\end{CodeInput}
\end{CodeChunk}

@@

\begin{center}\includegraphics{Imputation_of_Incomplete_Multilevel_Data_files/figure-latex/pop-ignored-1} \end{center}
